NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2019-11-13 Manuscript/manuscriptPGPE.tex

\subsection{Headgroup and glycerol backbone of POPE and POPG in MD simulations}
Headgroup and glycerol backbone order parameters of PE and PG lipids
show wide variation between different force fields
and none of the force fields reproduce all values within experimental error bars.
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
as observed previously also for PC and PS lipids \cite{botan15,antila19}.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{antila19}, but this is not the case for PE and PG lipids.
The poor performance of headgroup order parameters in Berger model can be probably explained by
ring like structures seen in Fig. 6 in Ref. \citenum{mukhopadhyay04}, which is a typical feature
for Berger based lipid force fields containing explicit hydrogen atoms in the head group \cite{zhao08,henin09,dahlberg10}.
The poor performance of glycerol backbone of Slipids simulations is systemically observed also
for other lipids in previous studies \cite{botan15,antila19}.
\todo{Should we comment more the relative quality of different force fields and/or make the subjective force field ranking figures?}

Without further discussion about poorly performing force fields, we focus on more detailed analysis
of CHARMM36 simulations, which give the best overall agreement with experiments for
the headgroup and glycerol backbone order parameters of all PC, PS, PG and PE lipids in this and our previous
studies \cite{botan15,antila19}. Even thought many order parameters in CHARMM36 simulations are
not within the experimental error bars, it seems to capture the essential differences between different headroups.
%(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}).
In previous study, CHARMM36 predicted the more negative $\beta$-carbon order parameter and larger forking of
the $\alpha$-carbon in PS headgroup than in PC \cite{antila19}.
In this work,
%the CHARMM36 simulations reproduce also the other essential differences in experimental
%headgroup order parameters between different headgroups (Fig. \ref{HGorderParameters}) despite the inaccuracies
%in individual segments:
the PE headgroup order parameters are similar to the PC \cite{botan15}
and $\beta$-carbon order parameters is positive in PG headgroup, while negative values observed in
other lipids. Therefore, we use CHARMM36 simulations to analyze the structural differences between
headgroups (Fig.~\ref{structures}).
\todo{The structures are already in Fig. \ref{structures}, but we need also the dihedral distributions to finish the discussion}.
%While rotation around N-C$_\alpha$-C$_\beta$-O$_\alpha$ dihedral was
%significantly different in PS and PC headgroups in previous work \cite{antila19}, essential differences
%between PC, PE and PG are not observed here (Fig.~\ref{structures}) \todo{We need also the dihedral distributions to finish this discussion.}. \\
%\todo{Why is the $\beta$-carbon order parameter of PG different to PC and PE then?}

\todo{Maybe we should analyze also some other structural features, at least P-N angles.}

@@

None of the force fields fully reproduces the PC headgroup order parameter response to the increasing amount of
PG, which may be related to the counterion binding affinity (see also the next section) \cite{antila19}.
In all force fields except Slipids, the order parameters of different
hydrogens attached to the $\alpha$-carbon are responsing differently when mixed with PE or PG lipids
\todo{Maybe we should figure out what is the reason for this?}.\\
\todo{Maybe we should analyze the P-N vector angle from different simulations?}.

>>>>

\subsection{Headgroup and glycerol backbone of POPE and POPG in MD simulations}
Headgroup and glycerol backbone order parameters of PE and PG lipids
show wide variation between different force fields
and none of the force fields reproduce all values within experimental error bars.
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
as observed previously also for PC and PS lipids \cite{botan15,antila19}.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{antila19}, but this is not the case for PE and PG lipids.
The poor performance of headgroup order parameters in Berger model can be probably explained by
ring like structures seen in Fig. 6 in Ref. \citenum{mukhopadhyay04}, which is a typical feature
for Berger based lipid force fields containing explicit hydrogen atoms in the head group \cite{zhao08,henin09,dahlberg10}.
The poor performance of glycerol backbone of Slipids simulations is systemically observed also
for other lipids in previous studies \cite{botan15,antila19}.
\todo{Should we comment more the relative quality of different force fields and/or make the subjective force field ranking figures?
https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/8}

Without further discussion about poorly performing force fields, we focus on more detailed analysis
of CHARMM36 simulations, which give the best overall agreement with experiments for
the headgroup and glycerol backbone order parameters of all PC, PS, PG and PE lipids in this and our previous
studies \cite{botan15,antila19}. Even thought many order parameters in CHARMM36 simulations are
not within the experimental error bars, it seems to capture the essential differences between different headroups.
%(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}).
In previous study, CHARMM36 predicted the more negative $\beta$-carbon order parameter and larger forking of
the $\alpha$-carbon in PS headgroup than in PC \cite{antila19}.
In this work,
%the CHARMM36 simulations reproduce also the other essential differences in experimental
%headgroup order parameters between different headgroups (Fig. \ref{HGorderParameters}) despite the inaccuracies
%in individual segments:
the PE headgroup order parameters are similar to the PC \cite{botan15}
and $\beta$-carbon order parameters is positive in PG headgroup, while negative values observed in
other lipids. Therefore, we use CHARMM36 simulations to analyze the structural differences between
headgroups (Fig.~\ref{structures}).
\todo{The structures are already in Fig. \ref{structures}, but we need also the dihedral distributions to finish the discussion.
Maybe we should analyze also some other structural features, at least P-N angles. More discussion in issue
https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/9}.
%While rotation around N-C$_\alpha$-C$_\beta$-O$_\alpha$ dihedral was
%significantly different in PS and PC headgroups in previous work \cite{antila19}, essential differences
%between PC, PE and PG are not observed here (Fig.~\ref{structures}) \todo{We need also the dihedral distributions to finish this discussion.}. \\
%\todo{Why is the $\beta$-carbon order parameter of PG different to PC and PE then?}

@@

None of the force fields fully reproduces the PC headgroup order parameter response to the increasing amount of
PG, which may be related to the counterion binding affinity (see also the next section) \cite{antila19}.
In all force fields except Slipids, the order parameters of different
hydrogens attached to the $\alpha$-carbon are responsing differently when mixed with PE or PG lipids
\todo{Maybe we should figure out what is the reason for this? \\
  Maybe we should analyze the P-N vector angle from different simulations? \\
  https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/10}.
